---
title: "Merge MMETSP and Genbank Data"
output: 
editor_options: 
  chunk_output_type: inline
---

# Setup
Load Libraries.
```{r}
library(tidyverse)
theme_set(theme_bw())
library("sf")
library("rnaturalearth")
library("rnaturalearthdata")
library(taxonomizr)
library(data.table)
```

# MMETSP

Read raw data.
```{r}
mmetsp_raw_data <- read_tsv('../data/sample-attr.tab.txt')
mmetsp_raw_data
```

Unpack attributes.
```{r}
mmetsp_wider <- mmetsp_raw_data %>%
  rename(sample_name_main = sample_name) %>%
  pivot_wider(id_cols = c('sample_id', 'sample_name_main'), names_from = "attr_type", values_from = "attr_value", names_repair = "unique")

mmetsp_wider
```

Some latitude and longitude entries have multiple coordinates.
```{r}
mmetsp_wider %>%
  mutate(latDup = ifelse(lengths(latitude) > 1, 1 , 0),
         longDup = ifelse(lengths(longitude) > 1, 1 , 0)) %>%
  filter(latDup == 1 | longDup == 1)
```

For duplicate lat/long, keep first latitude, and second longitude. Keep first sample name. For other columns with duplicate entries, paste together (comma separated).
```{r}
# keep first latitude, second longitude.
mmetsp_wider$longitude <- lapply(mmetsp_wider$longitude, function(x) if(length(x) > 1) x[[2]] else x)
mmetsp_wider$latitude <- lapply(mmetsp_wider$latitude, function(x) if(length(x) > 1) x[[1]] else x)

# keep first sample name
mmetsp_wider$sample_name <- lapply(mmetsp_wider$sample_name, function(x) if(length(x) > 1) x[[1]] else x)

# paste (comma separated) other duplicate entries.
mmetsp_wider$source_mat_id <- lapply(mmetsp_wider$source_mat_id, function(x) paste(x, collapse = ","))
mmetsp_wider$primary_citation <- lapply(mmetsp_wider$primary_citation, function(x) paste(x, collapse = ","))
mmetsp_wider$other_experimental_metadata_available <- lapply(mmetsp_wider$other_experimental_metadata_available, function(x) paste(x, collapse = ","))
mmetsp_wider$other_environmental_metadata_available <- lapply(mmetsp_wider$other_environmental_metadata_available, function(x) paste(x, collapse = ","))
mmetsp_wider$country <- lapply(mmetsp_wider$country, function(x) paste(x, collapse = ","))
mmetsp_wider$additional_citations <- lapply(mmetsp_wider$additional_citations, function(x) paste(x, collapse = ","))
```

Check that all list columns are of length one.
```{r}
mmetsp_wider %>%
  summarise(across(everything(), ~ max(lengths(.x)))) %>%
  t()
```
Unnest wide data, now that lists are only of length one.
```{r}
mmetsp_unnested <- mmetsp_wider %>%
  unnest()
```

```{r}
t(data.frame(as.list(colSums(!is.na(mmetsp_unnested))))) %>%
  as_tibble(rownames ="variable") %>%
  rename(count = V1) %>%
  arrange(-count)

```

```{r}
mmetsp_unnested %>%
  drop_na(depth)
```



# Mapping
```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")

coords_df <- mmetsp_unnested %>%
  select(latitude, longitude) %>%
  mutate_all(as.double) %>%
  na.omit() 
```

Map of MMETSP sampling points:
```{r}
ggplot(data = world) +
    geom_sf() +
    geom_point(data = coords_df, aes(x = longitude, y = latitude), size = 2, 
        shape = 23, fill = "red")+
  coord_sf(xlim = c(-180, 180), ylim = c(-90,90), expand = FALSE)

ggsave("../output/mmetsp_map.png")
```


Select and unnest taxon info.
```{r}
mmetsp_taxon <- mmetsp_wider %>%
  select(sample_id, sample_name_main, taxon_id, phylum, class, order, genus, species, strain, fastq_file, latitude, longitude) %>%
  unnest() %>%
  mutate(
    genus_species_strain = gsub(" ", "_", paste(genus, species, strain, sep = "_"))
  )
mmetsp_taxon
```


Select only barebones.
```{r}
mmetsp_select <- mmetsp_taxon %>%
  select('sample_id', 'sample_name_main', 'taxon_id', 'genus_species_strain', 'fastq_file')
```

# Genbank

Specify column names.
```{r}
colNames <- "assembly_accession, bioproject, biosample, wgs_master, refseq_category, taxid, species_taxid, organism_name, infraspecific_name, infraspecific_name2, isolateversion_status, assembly_level, release_type, genome_rep, seq_rel_date, asm_name, submitter, gbrs_paired_asm, paired_asm_comp, ftp_path, excluded_from_refseq, relation_to_type_material"
colNamesVec <- unlist(str_split(colNames, ", "))
```

Read in metadata for all Genbank genomes.
```{r}
# genome filename ending
suffix <- "_genomic.fna.gz"

genbank <- read_tsv('../data/assembly_summary_genbank.txt',
                    comment = "#",
                    col_names = colNamesVec) %>%
  # remove entries missing ftp genome path
  drop_na(ftp_path) %>%
  mutate(taxid = as.character(taxid),
         species_taxid = as.character(species_taxid)) %>%
  rowwise() %>%
  mutate(
    # add full filename and ftp path of genbank genomes
    genome_filename = paste(tail(str_split(ftp_path, '/')[[1]], 1), suffix, sep = ""),
    genome_ftp_path = paste(ftp_path, genome_filename, sep = "/")
  )
```

# Join MMETSP/Genbank by taxid
```{r}
paste("/iplant/home/shared/imicrobe/projects/104/samples", "sample", 
      paste("sampleid", ".fastq.tar", sep = ""), 
      sep = "/")
```


Join MMETSP and Genbank data by taxon id.
```{r}
# genome filename ending
suffix <- "_genomic.fna.gz"

# join mmetsp & genbank
join_taxid <- inner_join(mmetsp_unnested, genbank, by = c('taxon_id' = 'taxid')) %>%
  # remove duplicates
  distinct(sample_id, .keep_all = TRUE) %>%
  distinct(taxon_id, .keep_all = TRUE) %>%
  rowwise() %>%
  mutate(
    # add full filename and ftp path of genbank genomes
    genome_filename = paste(tail(str_split(ftp_path, '/')[[1]], 1), suffix, sep = ""),
    genome_ftp_path = paste(ftp_path, genome_filename, sep = "/"),
    strain = str_replace(strain, " ", "-"),
    new_fastq_ftp = paste("/iplant/home/shared/imicrobe/projects/104/samples", sample_id, 
      paste(sample_name_main, ".fastq.tar", sep = ""), 
      sep = "/"),
    genome_name = str_split(genome_filename, ".fna.gz")[[1]][1]
  ) 
  
join_taxid$new_fastq_ftp[1]
```




# Write taxid joined 
CSV of all info:


```{r}
write_csv(join_taxid, '../data/mmetsp_ncbi_genome_info.csv')
```

FTP paths:
```{r}
readr::write_lines(join_taxid$genome_ftp_path, '../data/genome_ftp_paths.txt')
```

# Get Genbank Taxonomy info
Download NCBI taxonomy SQL database (only need to do this once!).
```{r}
##prepareDatabase('../data/accessionTaxa.sql')
```

Select only genome info from genbank.
```{r}
genbank_select <- genbank %>%
  select(taxid, species_taxid, genome_ftp_path, infraspecific_name, infraspecific_name2, genome_filename) %>%
  ungroup() %>%
  #keep only unique taxid
  distinct(taxid, .keep_all = TRUE)
genbank_select
```

Get taxonomy info from taxid.
```{r}
genbankTaxonomy <- getTaxonomy(unique(genbank$taxid), sqlFile = "../data/accessionTaxa.sql") %>%
  data.frame(taxid = row.names(.), row.names = NULL) %>%
  # remove spaces in taxid
  mutate(taxid = str_trim(taxid))
```

Add taxonomy info to genbank df. 
```{r}
genbank_taxon_joined <- left_join(genbank_select, genbankTaxonomy, by = "taxid")
genbank_taxon_joined
```
# Match genbank and mmetsp taxonomy

Find matching taxid.
```{r}
taxid_match <- genbank_taxon_joined %>%
  filter(taxid %in% mmetsp_unnested$taxon_id | species_taxid %in% mmetsp_unnested$taxon_id)
```

Matching genus, not taxid.
```{r}
genus_match <- genbank_taxon_joined %>%
  drop_na(genus) %>%
  filter(genus %in% mmetsp_unnested$genus,
         !(taxid %in% mmetsp_unnested$taxon_id | species_taxid %in% mmetsp_unnested$taxon_id)) %>%
  group_by(genus) %>%
  top_n(1) %>%
  ungroup()
```

```{r}
taxid_match
```

```{r}
genus_match
```

Matching family, not lower.
```{r}
family_match <- genbank_taxon_joined %>%
  drop_na(family) %>%
  filter(family %in% mmetsp_unnested$family,
         !(genus %in% mmetsp_unnested$genus),
         !(taxid %in% mmetsp_unnested$taxon_id | species_taxid %in% mmetsp_unnested$taxon_id)) %>%
  group_by(family) %>%
  top_n(1)
```

Matching order, not lower.
```{r}
order_match <- genbank_taxon_joined %>%
  drop_na(order) %>%
  filter(order %in% mmetsp_unnested$order, 
         !(family %in% mmetsp_unnested$family),
         !(genus %in% mmetsp_unnested$genus),
         !(taxid %in% mmetsp_unnested$taxon_id | species_taxid %in% mmetsp_unnested$taxon_id)) %>%
  group_by(order) %>%
  top_n(1) %>%
  ungroup()
```

Matching class, not lower.
```{r}
class_match <- genbank_taxon_joined %>%
  drop_na(class) %>%
  filter(class %in% mmetsp_unnested$class, 
         !(order %in% mmetsp_unnested$order), 
         !(family %in% mmetsp_unnested$family),
         !(genus %in% mmetsp_unnested$genus),
         !(taxid %in% mmetsp_unnested$taxon_id | species_taxid %in% mmetsp_unnested$taxon_id)) %>%
  group_by(class) %>%
  top_n(1) %>%
  ungroup()
```

Combine all filtered taxonomy to master df.
```{r}
genbank_mmetsp <- rbind(class_match, order_match, family_match, genus_match, taxid_match) %>%
  arrange(species) %>%
  relocate(infraspecific_name, .after = species) %>%
  relocate(infraspecific_name2, .after = infraspecific_name)

genbank_mmetsp
```

